Model Search in Relational Domains

### 2016-10-22 deductions.tex
moved back from bitbucket to github

\section{Abstract Model Search}
%
\Omit {
\begin{figure}[t]
\scriptsize
\begin{tabular}{l|l|l}
\hline
C program & SSA & Octagon \\
\hline
\begin{lstlisting}[mathescape=true,language=C]
int main() {
 unsigned x, y;
 __CPROVER_assume(x==y);
 x++;
 assert(x==y+1);
}
\end{lstlisting}
&
\begin{minipage}{4.40cm}
$\begin{array}{l@{\,\,}c@{\,\,}l}
SSA &\iff& ((g0 == TRUE) \land \\
    &    & (cond == (x == y)) \land \\
    &    & (g1 == (cond \&\& guard0)) \land \\
    &    & (x' == 1u + x) \land \\
    &    & (x' == 1u + y || !1))
\end{array}$
\end{minipage}
&
\begin{minipage}{3.75cm}
$\begin{array}{l@{\,\,}c@{\,\,}l}
C &\iff& ((x' > 1) \land (-x'-y < -2) \land \\
  &    & (-x-x' < -2) \land (y-x' < 0) \land \\                                                                
  &    & (x-x' < 0) \land (y > 0) \land \\
  &    & (x > 0) \land (-x'-y < 0) \land \\
  &    & (x+y > 1) \land (y-x < 1) \land \\
  &    & (x'-y < 2) \land (x-y < 1) \land \\
  &    & (x+y > 0) \land (x+x' > 0) \land \\
  &    & (x'-x < 2))
\end{array}$
\end{minipage}
\\
\hline
\end{tabular}
\caption{C Program, corresponding SSA and Octagonal Inequalities}
\label{ssa}
\end{figure}
}
%
%    
\begin{algorithm2e}[t]
\DontPrintSemicolon
\SetKw{return}{return}
\SetKwData{sat}{sat}
\SetKwData{conflict}{conflict}
\SetKwData{unsat}{unsat}
\SetKwData{unknown}{unknown}
\SetKwData{true}{true}
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKwFor{Loop}{Loop}{}{}
\SetKw{KwNot}{not}
\begin{small}
\Input{A program in the form of a set of abstract transformers $\abstransset$,
a propagation trail $\trail$, and a reason trail $\reasons$.}
\Output{\sat or \conflict or \unknown}
$\worklist \leftarrow \initworklist_{\propheur}(\abstransset)$ \;
\While{$!\mathit{worklist.empty}()$} 
{
%  $(\abstransel,\livevars) \leftarrow \mathit{worklist.pop}()$ \;
  $\abstransel{\subdomain} \leftarrow \mathit{worklist.pop}()$ \;
%  $v \leftarrow ded(s, \makesubdomain(\livevars), \abs(\trail))$\;
  $v \leftarrow \abstransel{\subdomain}(\abs(\trail))$\;
  \uIf{$v = \bot$} {
    $\reasons[\bot] \leftarrow \abstransel{\subdomain}$ \;
%\pscmt{how is this used in the conflict
%    analysis? wouldn't it be more uniform to add $\bot$ to the $\trail$?}
%    \rmcmt{UIP uses the transformer stored at $\reasons[\bot]$ to derive the
%    cut. No, we dont want to make the trail inconsistent by adding $\bot$}
    $\mathit{worklist.clear}()$ \;
%\pscmt{why can't you clean up the worklist during conflict analysis? it seems inefficient to throw everything away...}
%    \rmcmt{When a bottom is deduced, the direction of search
%    changes--so the worklist is populated with new set of statement and must be
%    flushed out} 
    \return \conflict \;
  }
  \uElse
  {
    $\trail \leftarrow \trail \cdot \decomp(v)$ \; 
%    $v \leftarrow v \meet \mathit{meet\_irrd}$ \; 
    $\reasons[|\trail|] \leftarrow s$ \;
    $\updateworklist_{\propheur}(v, \worklist)$ \; 
  }
}
\lIf{$\abs(\trail)$ is $\gamma$-complete} {
  \return \sat;
}
 \return \unknown \;

@@

To make our algorithm efficient, we have to focus abstract
transformers on performing only the minimally necessary
work. 
%
We thus define a specialised variant of the transformer to compute
deductions w.r.t.\ a given subdomain $\subdomain\subseteq \domain$.
A subdomain contains a chosen subset of the elements in $\domain$ including $\bot$ and $\top$.
\[\abstrans{\domain}{\constraint}^\subdomain(\absval)=\absval\meet_\domain \alpha_\subdomain(\{\val\mid \val\models \constraint\})\]
For $\subdomain=\domain$, the abstract transformer behaves in the same way as the one defined in Section~\ref{sec:domains}.
%
The restriction to a subdomain makes it less precise. Therefore, we have the property $\abstrans{\domain}{\constraint}^\domain(\absval)\sqsubseteq \abstrans{\domain}{\constraint}^\subdomain(\absval)$.
%
\pscmt{explain $\makesubdomain(Vars)$}
For example, for
$\absval=(0\leq y \leq 1 \wedge 5\leq z)$ we have
$\abstrans{\domain}{x=y+z}^{\makesubdomain(\{x\})}(\absval)=(x\geq 6)$, whereas
$\abstrans{\domain}{x=y+z}^{\makesubdomain(\{y,z\})}=\top$.
%
The advantage of this definition is that with the help of $\subdomain$
we can now control how we propagate within our constraint system.
%
For the example above, $\subdomain=\makesubdomain(\{x\})$ performs a
right-hand side to left-hand side propagation and hence emulates a
forward analysis on the input program; whereas with
$\subdomain=\makesubdomain(\{y,z\})$) we emulate a backward analysis.
%
We call the propagation with $\subdomain=\domain$ \emph{multi-way
  propagation}, which simultaneously performs forward and backward
propagation.

@@

\pscmt{give an example for $ded$, link it to the reason trail}

\pscmt{define $ded$ for all statements, connect this to worklist}

\pscmt{link livevars to $L$ -- the subdomain spanned by livevars}

%-------------------------------------------------------------------------------
\subsection{Algorithm for Deduction Phase}
%-------------------------------------------------------------------------------
%
\rmcmt{Talk about lazy closure operation}

\pscmt{
Rough and sketchy:
%
$espan(V)=$all meet irreducibles $\in A$ that contain at most one variable that is not in $V$.
%
$select(S,PV)$ is the set of statements that contain variables in $PV$ where $PV$ is the set of variables in the previous deduction.
%
$LV(s)$ are the live variables of a statement $s$, i.e.\ the intersection of $PV$ with the variables in $s$.
\begin{theorem}
$$
\gfp\lambda a. \bigsqcap_{s\in S} ded(s,a,A) = 
\gfp\lambda a. \bigsqcap_{s\in select(S,PV)} ded(s,a,espan(LV(s)))
$$
\end{theorem}
Proof sketch: see Appendix
}

Algorithm~\ref{Alg:ms} presents the deduction phase in abstract model
search procedure. \pscmt{The propagation heuristics $H$ provides two
  functions $\initworklist$ and
  $\updateworklist$. Describe them for the different
  heuristics (already done for initialize, TBD for update...}
A {\em worklist} is a queue which contains \rmcmt{abstract deduction
transformers}.  A {\em worklist} is initialized following a heuristics 
choice for propagation strategies. A {\em forward} propagation strategy 
initializes the worklist with \rmcmt{abstract} transformers that 
has constants in the right-hand side. Whereas, a {\em backward} 
iteration strategy starts with a set of assertions. On the other hand, a 
{\em multi-way} iteration strategy initializes the worklist with set of all 
transformers and performs both forward and backward propagations in tandem.  
For every meet irreducible deduction transformer \pscmt{rename} $s \in S$ 
passed to the abstract domain, if the domain infers a new meet irreducible 
$\mathit{meet\_irrd}$, it is added to $\trail$ \pscmt{It usually infers several ones.}.  Additionally, the reason 
trail $\reasons$ \pscmt{define} is updated with the transformer $s$ as the reason for inferring
$\mathit{meet\_irrd}$. If a conflict is deduced, that is, $\mathit{meet\_irrd} = \bot$ 
is inferred by the domain, the algorithm terminates with \textsf{conflict}.  However,
when a fixed-point is reached, the abstract valuation $v$ is checked for set of
models.  This is determined by checking whether the set of assignments in $v$ is 
$\gamma$-complete~\cite{dhk2013-popl}. If $v$ is $\gamma$-complete, that is, the 
abstract valuation $v$ is abstractly satisfying, the abstract model 
search terminates with \textsf{sat}.  Otherwise, the algorithm returns \textsf{unknown} and the 
model search procedure makes a new decision.

>>>>

\section{Abstract Model Search}
%
\Omit {
\begin{figure}[t]
\scriptsize
\begin{tabular}{l|l|l}
\hline
C program & SSA & Octagon \\
\hline
\begin{lstlisting}[mathescape=true,language=C]
int main() {
 unsigned x, y;
 __CPROVER_assume(x==y);
 x++;
 assert(x==y+1);
}
\end{lstlisting}
&
\begin{minipage}{4.40cm}
$\begin{array}{l@{\,\,}c@{\,\,}l}
SSA &\iff& ((g0 == TRUE) \land \\
    &    & (cond == (x == y)) \land \\
    &    & (g1 == (cond \&\& guard0)) \land \\
    &    & (x' == 1u + x) \land \\
    &    & (x' == 1u + y || !1))
\end{array}$
\end{minipage}
&
\begin{minipage}{3.75cm}
$\begin{array}{l@{\,\,}c@{\,\,}l}
C &\iff& ((x' > 1) \land (-x'-y < -2) \land \\
  &    & (-x-x' < -2) \land (y-x' < 0) \land \\                                                                
  &    & (x-x' < 0) \land (y > 0) \land \\
  &    & (x > 0) \land (-x'-y < 0) \land \\
  &    & (x+y > 1) \land (y-x < 1) \land \\
  &    & (x'-y < 2) \land (x-y < 1) \land \\
  &    & (x+y > 0) \land (x+x' > 0) \land \\
  &    & (x'-x < 2))
\end{array}$
\end{minipage}
\\
\hline
\end{tabular}
\caption{C Program, corresponding SSA and Octagonal Inequalities}
\label{ssa}
\end{figure}
}
%
%    
\begin{algorithm2e}[t]
\DontPrintSemicolon
\SetKw{return}{return}
\SetKwData{sat}{sat}
\SetKwData{conflict}{conflict}
\SetKwData{unsat}{unsat}
\SetKwData{unknown}{unknown}
\SetKwData{true}{true}
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKwFor{Loop}{Loop}{}{}
\SetKw{KwNot}{not}
\begin{small}
\Input{A program in the form of a set of abstract transformers $\abstransset$,
a propagation trail $\trail$, and a reason trail $\reasons$.}
\Output{\sat or \conflict or \unknown}
$\worklist \leftarrow \initworklist_{\propheur}(\abstransset)$ \;
\While{$!\mathit{worklist.empty}()$} 
{
  $\abstransel{\subdomain} \leftarrow \mathit{worklist.pop}()$ \; 
  \rmcmt{how do we get the $\subdomain$ at this point ?}
  $\newdeductions \leftarrow \onlynew(\abstransel{\subdomain}(\abs(\trail)))$\;
  \uIf{$\newdeductions = \bot$} {
    $\reasons[\bot] \leftarrow \abstransel{\subdomain}$ \rmcmt{shouldn't this be
    just $\abstransset$} \;
    $\mathit{worklist.clear}()$ \;
    \return \conflict \;
  }
  \uElse
  {
%    $\newdeductions=\decomp(\absval)\setminus\decomp(\abs(\trail))$\;
    $\trail \leftarrow \trail \cdot \decomp(\newdeductions)$ \; 
    $\reasons[|\trail|] \leftarrow \abstransel{\subdomain}$ \;
    $\updateworklist_{\propheur}(\newdeductions, \worklist)$ \; 
  }
}
\lIf{$\abs(\trail)$ is $\gamma$-complete} {
  \return \sat;
}
 \return \unknown \;

@@

To make our algorithm efficient, we have to focus abstract
transformers on performing only the minimally necessary
work. 
%
We thus define a specialised variant of the abstract transformer to compute
deductions w.r.t.\ a given subdomain $\subdomain\subseteq \domain$.
A subdomain contains a chosen subset of the elements in $\domain$ including $\bot$ and $\top$
\begin{equation}\label{eq:at2}
\abstrans{\domain}{\constraint}^\subdomain(\absval)=\absval\meet_\domain \alpha_\subdomain(\{\val\mid \val\in\gamma_\domain(\absval), \val\models \constraint\})
\end{equation}
For $\subdomain=\domain$, the abstract transformer behaves in the same way as the one defined in Section~\ref{sec:domains}.

%We construct the subdomain from a set of variables $\vars$ such that $\subdomain_\domain(\vars)$ is the set of meet irreducibles $\in \domain$ that contain at last one variable in $\vars$ and most one variable that is not in $\vars$. \pscmt{split into two parts}
%
%For example, $\subdomain=\domain[\{y\}]$ being the octagons domain over variables $\{x,y,z\}$ contains all octagonal meet irreducibles involving $\{y\}$, and the pairs $\{x,y\}$ and $\{y,z\}$, but not the meet irreducibles\pscmt{wrong} over $\{x\}$, $\{z\}$ and~$\{x,z\}$.

The first advantage of this definition is that with the help of $\subdomain$
we can control how we perform propagation using our abstract transformers.
%
For example, for $\absval=(0\leq y \leq 1 \wedge 5\leq z)$, we have
$\abstrans{\intervals[\{x,y,z\}]}{x=y+z}^{\intervals[\{x\}]}(\absval)=\absval\meet(x\geq
6)$. For $x=y+z$, this performs a right-hand side to left-hand side propagation and
hence emulates a forward analysis, whereas
$\abstrans{\intervals[\{x,y,z\}]}{x=y+z}^{\intervals[\{y,z\}]}=\absval\meet\top$
emulates a backward analysis.

We call the propagation with arbitrary subdomains,
e.g. $\subdomain=\domain$, \emph{multi-way propagation}, which is able
to simultaneously perform forward and backward propagation.
%
Note that the restriction to a subdomain makes the transformer less
precise. Therefore, we have the property
$\abstrans{\domain}{\constraint}^\domain(\absval)\sqsubseteq
\abstrans{\domain}{\constraint}^\subdomain(\absval)$.
However, the choice of a subdomain makes the deductions more efficient.

The second advantage of Eq.~(\ref{eq:at2}) is that we can compute the
\emph{closure} operation for relational domains in a lazy manner. To
this end, we construct a $\subdomain=\makesubdomain_\domain(\vars)$
which is sufficient to perform one step of the closure operation each
time the abstract transformer is applied.
%
For example, let us consider $\domain=\octagons[\{x,y,z\}]$ and
$\vars=\{y\}$. \rmcmt{A binary} octagon \rmcmt{element} relate at 
most two variables. Thus it is sufficient to consider the subdomain
$\makesubdomain_\domain(\{y\})=\octagons[\{y\}]\cup\octagons[\{x,y\}]\cup\octagons[\{y,z\}]$,
which will compute the one-step transitive relations of $y$ with each
of the other variables. If this indeed affects $x$ and $z$ then their
relations will be inferred in a later step of the propagation
algorithm, thus gradually computing the closure. Note that this might
increase the number of propagation steps, but it reduces the size of
the domain used in the abstract transformer.

@@

%-------------------------------------------------------------------------------
\subsection{Algorithm for Deduction Phase}
%-------------------------------------------------------------------------------
%

Algorithm~\ref{Alg:ms} presents the deduction phase $\deduce$ in 
abstract model search procedure.  The input to the algorithm is 
the set of abstract transformers, a propagation trail and a reason 
trail.  Additionally, the procedure $\deduce$ is parameterised with 
a propagation heuristic.  The propagation heuristics provides two 
functions $\initworklist$ and $\updateworklist$.  Given a set of 
variables in the \rmcmt{constraint system}, a subdomain is essentially a 
subset of variables \rmcmt{how we are defining subdomain?} that guides the propagations in forward, 
backward or multi-way.  For example, the subdomain consist of variables 
in lhs in forward propagation heuristic.  For backward, the subdomain 
consist of variables in rhs, whereas the subdomain contains all variables 
for multi-way propagation.  A {\em worklist} is a queue which contains 
a set of abstract deduction transformers.  
The initWorklist initializes the worklist 
following a heuristics choice for propagation strategies.  A {\em forward} 
propagation strategy initializes the worklist with abstract deduction 
transformers that contains constants in the right-hand side. Whereas, a 
{\em backward} iteration strategy starts with a set of assertions. On the 
other hand, a {\em multi-way} iteration strategy initializes the worklist 
with set of all transformers and performs both forward and backward 
propagations in tandem.  \rmcmt{wrong to say abstract deduction transformer that 
has lhs, rhs since since it also includes the live variables}.

Given a constraint system and a domain, an updateWorklist function adds 
new abstract deduction transformers to the worklist for those variables in 
the subdomain that appear in newDeductions.  It also updates the subdomain of 
existing abstract deduction transformers that are in the worklist.

The function
$\onlynew(\absval)=\bigsqcap(\decomp(\absval)\setminus\decomp(\abs(\trail)))$
filters out all known meet irreducibles in the abstract value.

For every abstract deduction transformer in the worklist, 
the set of new meet irreducibles inferred by the domain is added 
to the $\trail$.  Additionally, the reason trail $\reasons$ is 
updated with the corresponding abstract transformer as the reason for inferring 
those new meet irreducibles.  If a conflict is deduced, 
%that is, $\mathit{meet\_irrd} = \bot$ is inferred by the domain, 
the algorithm terminates with \textsf{conflict}.  However,
when a fixed-point is reached, the abstract value $v$ computed from 
the trail is checked for set of models.  This is determined by 
checking whether the set of assignments in $v$ is 
$\gamma$-complete~\cite{dhk2013-popl}. If $v$ is $\gamma$-complete, that is, the 
abstract valuation $v$ is abstractly satisfying, the abstract model search 
terminates with \textsf{sat}.  Otherwise, the algorithm returns \textsf{unknown} and the 
model search procedure makes a new decision.
